---
title: "R Coding Exercise"
editor: 
  markdown: 
    wrap: 72
---

::: r-coding-exercise-custom-special-style
The analysis indicates that there is no significant correlation between
population size and life expectancy among African countries for the year
2000. This suggests that other factors not included in this model might
be more influential in determining life expectancy.

# Loading required libraries

# `dslabs` is used to access the gapminder dataset which includes global demographic data

```{r}
library("dslabs")
```
:::

2.  Accessing and Exploring the Dataset Once the necessary package is
    loaded, the next step is to explore the dataset you'll be working
    with. Proper comments explain both what the code does and why you
    are doing it.

# Accessing the documentation for the Gapminder dataset

# This helps understand the structure and variables included in the dataset

```{r}
help(gapminder)
```

# Exploring the structure of the gapminder dataset

# `str()` provides a concise summary of the dataset structure showing types of variables and sample data

```{r}
str(gapminder)
```

# Summarizing the gapminder dataset

# `summary()` provides a statistical summary for each variable, useful for initial data assessment

```{r}
summary(gapminder)
```

# Checking the data type of the gapminder dataset

# Ensuring that the dataset is a dataframe which is suitable for data manipulation tasks

```{r}
class(gapminder)
```

3.  Data Manipulation: When manipulating data, clearly comment on each
    step to explain the purpose and the operations being performed.

# Filtering data to include only African countries

# Creating a new data object `africadata` to specifically analyze this continent

```{r}
africadata <- gapminder[gapminder$continent == "Africa", ]
```

# Checking the structure of the filtered data

```{r}
str(africadata)
```

# Summarizing the African data subset:

```{r}
summary(africadata)
```

4.  Creating Specific Data Views: Creating specific views of the data
    should be clearly commented to explain what each view represents.

# Creating a data object with selected variables:

We’ll create two new objects: one containing only infant_mortality and
life_expectancy, and the other containing population and
life_expectancy.

```{r}
health_data <- africadata[, c("infant_mortality", "life_expectancy")]
```

# Creating another view focusing on population size and life expectancy

```{r}
population_data <- africadata[, c("population", "life_expectancy")]
```

# Exploring the structure and summary of these new datasets

```{r}
str(health_data)
summary(health_data)
str(population_data)
summary(population_data)
```

5.  Plotting and Analysis: Documenting plotting and further analysis
    helps in understanding the purpose and interpretation of
    visualizations and statistical results.

# Plotting life expectancy vs infant mortality:

# Expecting to see a negative correlation indicating higher infant mortality associated with lower life expectancy

```{r}
plot(africadata$infant_mortality, africadata$life_expectancy, main="Life Expectancy vs Infant Mortality", xlab="Infant Mortality", ylab="Life Expectancy")
```

# Plotting life expectancy vs population size:

# Using log scale for population to better visualize data across wide range of values

```{r}
plot(africadata$population, africadata$life_expectancy, main="Life Expectancy vs Population Size", xlab="Log of Population Size", ylab="Life Expectancy",log="x")
```

#More Data Processing: 1.Load necessary library:

```{r}
library(dplyr)
```

#1. Identify years with missing data for infant mortality

```{r}
missing_data_by_year <- africadata %>%
  group_by(year) %>%
  summarize(missing_infant_mortality = sum(is.na(infant_mortality)))
```

# Calculate total missing values for infant mortality:

```{r}
total_missing_infant_mortality <- sum(is.na(africadata$infant_mortality))
print(paste("Total missing values for infant mortality:", total_missing_infant_mortality))
```

# Print years with missing data

```{r}
print("Years with missing data for infant mortality:")
print(missing_data_by_year)
```

# 2. Filter data for the year 2000:

```{r}
data_2000 <- africadata %>%
  filter(year == 2000)
```

# 3. Explore the structure and summary of the dataset for the year 2000:

```{r}
print("Structure of data for the year 2000:")
str(data_2000)

print("Summary of data for the year 2000:")
summary(data_2000)
```

#More Plotting: \# Plot 1: Life Expectancy vs. Infant Mortality:

```{r}

plot(data_2000$infant_mortality, data_2000$life_expectancy,
     main = "Life Expectancy vs. Infant Mortality (2000)",
     xlab = "Infant Mortality",
     ylab = "Life Expectancy",
     pch = 19,col='blue')  # pch = 19 makes the points solid circles

```

# Plot 2: Life Expectancy vs. Population Size

```{r}
plot(data_2000$population, data_2000$life_expectancy,
     main = "Life Expectancy vs. Population Size (2000)",
     xlab = "Population Size",
     ylab = "Life Expectancy",
     log = "x",  # sets the x-axis to logarithmic scale
     pch = 19, col='red')  # pch = 19 makes the points solid circles
```

# Summary of model fits:

# Using linear models to understand the influence of infant mortality and population size on life expectancy

# Linear Regression for Life Expectancy vs. Infant Mortality:

```{r}
fit1 <- lm(life_expectancy ~ infant_mortality, data = data_2000)
summary(fit1)  # Outputs the summary of the regression model
```

## Regression Analysis: Life Expectancy vs. Infant Mortality (2000)

### Objective

The goal was to analyze the impact of infant mortality on life
expectancy among African countries in the year 2000.

### Methodology

A linear regression model was fitted with life expectancy as the
dependent variable and infant mortality as the predictor.

### Results

-   **Model Coefficients**:
    -   **Intercept**: The model estimates a life expectancy of 71.29
        years when the infant mortality rate is zero.
    -   **Infant Mortality Coefficient**: The negative coefficient of
        -0.18916 indicates that as infant mortality increases by one
        unit, life expectancy decreases by approximately 0.189 years.
-   **Statistical Significance**:
    -   The p-value for the infant mortality coefficient is extremely
        low (\< 2.83e-08), suggesting a statistically significant
        negative impact of infant mortality on life expectancy.
-   **Model Fit**:
    -   The R-squared value of 0.4701 implies that about 47.01% of the
        variability in life expectancy across the dataset can be
        explained by differences in infant mortality rates.
    -   The residual standard error of 6.221 indicates the average
        deviation of the observed life expectancies from the values
        predicted by the model.

### Conclusion

The analysis confirms a significant negative correlation between infant
mortality and life expectancy for the year 2000 in African countries.
This suggests that policies aimed at reducing infant mortality could be
effective in increasing life expectancy.

# Linear Regression for Life Expectancy vs. Population Size

# Using logarithmic transformation for population size:

```{r}
fit2 <- lm(life_expectancy ~ (population), data = data_2000)
summary(fit2)  # Outputs the summary of the regression model
```

## Regression Analysis: Life Expectancy vs. Population Size (2000)

### Objective

The aim was to explore whether the size of a population in African
countries in the year 2000 impacts life expectancy.

### Methodology

A linear regression model was employed with life expectancy as the
outcome and the population size as the predictor.

### Results

-   **Model Coefficients**:
    -   **Intercept**: The model predicts a base life expectancy of 5593
        years, adjusting for population size.
    -   **Log(Population) Coefficient**: The coefficient of 2.756e-08
        suggests a negligible change in life expectancy with increasing
        population size.
-   **Statistical Significance**:
    -   The p-value for the log of population size is 0.6159, indicating
        that this model does not find a statistically significant impact
        of population size on life expectancy.
-   **Model Fit**:
    -   The R-squared value is only 0.005176, highlighting that
        population size explains a very small fraction of the
        variability in life expectancy.
    -   The negative adjusted R-squared value suggests that the model
        does not adequately predict life expectancy and might perform
        worse than a model with no predictors at all.

### Conclusion

The analysis indicates that there is no significant correlation between
population size and life expectancy among African countries for the year
2000. This suggests that other factors not included in this model might
be more influential in determining life expectancy.
