---
title: "Exploratory Data Analysis of NCHS Data"
editor: 
  markdown: 
    wrap: 72
---

<div class="r-coding-exercise-custom-special-style">

In this analysis, we will explore a dataset from the National Center for
Health Statistics (NCHS) that contains information about death rates and
life expectancy at birth. Our goal is to clean the data, perform
exploratory data analysis (EDA), and summarize the variables by their
distributions.

# Load necessary libraries:

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(here)
```

# Read the data:

```{r}
# Read the data
file_path <- here("cdcdata-exercise","NCHS_-_Death_rates_and_life_expectancy_at_birth_20240701.csv")
data <- read_csv(file_path)
```

# Display the first few rows of the dataset:

```{r}
head(data)
```

# Check the structure of the dataset:

```{r}
str(data)
```

# Clean the data:

# Remove any rows with NA values:

```{r}
data <- data %>%
  filter_all(any_vars(!is.na(.)))
```

# Replace missing values coded as 999 with NA:

```{r}
data <- data %>%
  mutate_if(is.numeric, ~na_if(., 999))
```

# Select a few variables for analysis (adjust based on actual variable names in the dataset):

```{r}
selected_data <- data %>%
  select(Year, Sex, Race, "Average Life Expectancy (Years)", "Age-adjusted Death Rate")
```

# Exploratory Data Analysis (EDA):

# Summary statistics:

```{r}
summary(selected_data)
```

## Summary of categorical variables:

We'll start by summarizing the categorical variables (Sex and Race) by
showing the percentage of each category.

```{r}
cat_summary_sex <- selected_data %>%
  group_by(Sex) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

cat_summary_race <- selected_data %>%
  group_by(Race) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

print(cat_summary_sex)
print(cat_summary_race)
```

Sex Distribution: The dataset is evenly distributed across the sex
categories: Both Sexes, Female, and Male. Each category represents
33.33% of the total dataset with 357 entries each. This balanced
representation ensures an unbiased analysis across sexes.

Race Distribution: The dataset is equally divided among the race
categories: All Races, Black, and White, with each category comprising
33.33% of the total dataset (357 entries each). This equal distribution
ensures comprehensive analysis across races.

# Distribution of categorical variables:

```{r}
cat_summary <- selected_data %>%
  group_by(Sex, Race) %>%
  summarize(count = n(), .groups = 'drop') %>%
  mutate(percentage = count / sum(count) * 100)
cat_summary
```

# Continuous Variables:

Next, we'll summarize the continuous variables (Age-adjusted Death Rate
and Average Life Expectancy (Years)) by plotting their distributions and
calculating their mean and standard deviation.

# Plot distribution of age-adjusted death rate:

```{r}
ggplot(selected_data, aes(x = `Age-adjusted Death Rate`)) +
  geom_histogram(binwidth = 50, fill = "blue", color = "black") +
  labs(title = "Distribution of Age-adjusted Death Rates", x = "Age-adjusted Death Rate", y = "Frequency")
```

The histogram shows age-adjusted death rates ranging from 0 to 4000,
with the highest frequencies. The frequency decreases as the death rate
increases, indicating fewer occurrences at higher death rates.This
visualization helps identify the concentration and spread of death rates
within the population, highlighting areas with the most common and least
common death rates.

# Summary statistics for age-adjusted death rate:

```{r}
death_rate_summary <- selected_data %>%
  summarize(mean = mean(`Age-adjusted Death Rate`, na.rm = TRUE),
            sd = sd(`Age-adjusted Death Rate`, na.rm = TRUE))

print(death_rate_summary)
```

The summary statistics table provides the mean and standard deviation
for the age-adjusted death rates. The mean age-adjusted death rate is
approximately 1593.06, with a standard deviation of 682.37. This
indicates that the average death rate is around 1593 deaths per 100,000
population, with considerable variability across the dataset.

# Plot distribution of Average life expectancy(Years):

```{r}
ggplot(selected_data, aes(x = `Average Life Expectancy (Years)`)) +
  geom_histogram(binwidth = 1, fill = "green", color = "black") +
  labs(title = "Distribution of Life Expectancy", x = "Life Expectancy (Years)", y = "Frequency")
```

The histogram illustrates the distribution of average life expectancy in
the dataset, ranging from approximately 30 to 85 years. The frequency
peaks around 70 years, indicating a higher number of occurrences in this
range. This visualization helps identify the spread and concentration of
life expectancy across the population, highlighting areas with the
highest and lowest frequencies.

# Summary statistics for Average life expectancy:

```{r}
life_expectancy_summary <- selected_data %>%
  summarize(mean = mean(`Average Life Expectancy (Years)`, na.rm = TRUE),
            sd = sd(`Average Life Expectancy (Years)`, na.rm = TRUE))

print(life_expectancy_summary)
```

The summary statistics table for average life expectancy shows a mean of
64.50 years and a standard deviation of 11.84 years. This indicates that
the average life expectancy in the dataset is approximately 64.5 years,
with a variability of about 11.8 years across the data points.

# Boxplots:

We'll also create boxplots to visualize the distribution of Age-adjusted
Death Rate by Sex and Average Life Expectancy (Years) by Race.

# Boxplot of age-adjusted death rate by sex:

```{r}
ggplot(selected_data, aes(x = Sex, y = `Age-adjusted Death Rate`, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Boxplot of Age-adjusted Death Rate by Sex", x = "Sex", y = "Age-adjusted Death Rate")
```

The boxplot shows the distribution of age-adjusted death rates across
different sex categories: Both Sexes, Female, and Male.

1.The median death rate for Both Sexes and Females is lower compared to
Males.

2.The interquartile range (IQR) is similar for Both Sexes and Females
but slightly larger for Males, indicating more variability in the male
category.

3.There are several outliers in the Male category, suggesting that there
are some instances of unusually high death rates.

This visualization highlights the differences in death rates among the
sex categories, with males showing higher median and more variability in
death rates.

# Boxplot of life expectancy by race:

```{r}
ggplot(selected_data, aes(x = Race, y = `Average Life Expectancy (Years)`, fill = Race)) +
  geom_boxplot() +
  labs(title = "Boxplot of Average Life Expectancy by Race", x = "Race", y = "Average Life Expectancy (Years)")
```

The boxplot compares average life expectancy across race categories: All
Races, Black, and White.

1.The median life expectancy is similar for All Races and White
categories, slightly lower for Black.

2.The interquartile range is largest for the White category, indicating
more variability.

3.The Black category shows the lowest median.

This visualization highlights the differences in life expectancy across
races, with the White category exhibiting the highest variability.

# **This section contributed by Seth Harris.**

### Load necessary libraries:

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(here)
library(synthpop)
```

### Generate synthetic data:

```{r}
# Load Sri's cleaned and processed data
file_path <- here("cdcdata-exercise", "NCHS_-_Death_rates_and_life_expectancy_at_birth_20240701.csv")
original_data <- read_csv(file_path)

# Rename columns to remove spaces
original_data <- original_data %>%
  rename(Average_Life_Expectancy_Years = `Average Life Expectancy (Years)`,
         Age_adjusted_Death_Rate = `Age-adjusted Death Rate`)

# Generate synthetic data using synthpop
set.seed(123)
synthetic_data <- syn(original_data)$syn
```

### Display the first few rows of the synthetic dataset:

```{r}
head(synthetic_data)
```

### Check the structure of the synthetic dataset:

```{r}
str(synthetic_data)
```

# Exploratory Data Analysis (EDA) on Synthetic Data

**In this section, I will perform some exploratory/descriptive analysis
on the synthetic dataset, similar to Sri's analysis.**

## **Summary statistics:**

```{r}
summary(synthetic_data)
```

## Summary of categorical variables:

I will summarize the categorical variables (e.g., `Sex` and `Race`) by
showing the percentage of each category.

```{r}
cat_summary_sex <- synthetic_data %>%
  group_by(Sex) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

cat_summary_race <- synthetic_data %>%
  group_by(Race) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

print(cat_summary_sex)
print(cat_summary_race)
```

## Distribution of categorical variables:

```{r}
cat_summary <- synthetic_data %>%
  group_by(Sex, Race) %>%
  summarize(count = n(), .groups = 'drop') %>%
  mutate(percentage = count / sum(count) * 100)
cat_summary
```

## Continuous Variables:

Next, I will summarize the continuous variables (Age_adjusted_Death_Rate
and Average_Life_Expectancy_Years) by plotting their distributions and
calculating their mean and standard deviation.

## Plot distribution of Age_adjusted_Death_Rate:

```{r}
ggplot(synthetic_data, aes(x = Age_adjusted_Death_Rate)) +
  geom_histogram(binwidth = 50, fill = "red", color = "orange") +
  labs(title = "Distribution of Age-adjusted Death Rates (Synthetic Data)", x = "Age-adjusted Death Rate", y = "Frequency")
```

### Distribution of Age-adjusted Death Rates (Synthetic Data)

#### Description:

This histogram shows the distribution of age-adjusted death rates in the
synthetic dataset. The x-axis represents the age-adjusted death rates,
while the y-axis shows the frequency of these rates.

#### Findings:

-   The histogram displays a right-skewed distribution, with a peak
    around the lower values of death rates.

-   Most of the synthetic data points have death rates concentrated
    between 500 to 1500, with decreasing frequency as the death rates
    increase.

-   The synthetic data mirrors the distribution characteristics observed
    in the original dataset, maintaining a similar skewness and
    concentration of values.

## Summary statistics for `Age_adjusted_Death_Rate`:

```{r}
death_rate_summary <- synthetic_data %>%
  summarize(mean = mean(Age_adjusted_Death_Rate, na.rm = TRUE),
            sd = sd(Age_adjusted_Death_Rate, na.rm = TRUE))

print(death_rate_summary)
```

## Plot distribution of `Average_Life_Expectancy_Years`:

```{r}
ggplot(synthetic_data, aes(x = Average_Life_Expectancy_Years)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "lightblue") +
  labs(title = "Distribution of Average Life Expectancy (Years) (Synthetic Data)", x = "Average Life Expectancy (Years)", y = "Frequency")
```

### Distribution of Average Life Expectancy (Years) (Synthetic Data)

#### Description:

This histogram illustrates the distribution of average life expectancy
in the synthetic dataset. The x-axis represents the average life
expectancy in years, and the y-axis shows the frequency of these values.

#### Findings:

-   The histogram reveals a distribution with a peak around 70 years,
    indicating a higher frequency of data points in this range.

-   The distribution is somewhat bimodal, with another smaller peak
    around 60 years.

-   The synthetic data closely follows the distribution observed in the
    original dataset, with a similar concentration of values around 70
    years and a similar spread of life expectancy values.

## Summary statistics for `Average_Life_Expectancy_Years`:

```{r}
life_expectancy_summary <- synthetic_data %>%
  summarize(mean = mean(Average_Life_Expectancy_Years, na.rm = TRUE),
            sd = sd(Average_Life_Expectancy_Years, na.rm = TRUE))

print(life_expectancy_summary)
```

## Boxplots:

I will also create boxplots to visualize the distribution of
Age_adjusted_Death_Rate by Sex and Average_Life_Expectancy_Years by
Race.

## Boxplot of `Age_adjusted_Death_Rate` by `Sex`:

```{r}
ggplot(synthetic_data, aes(x = Sex, y = Age_adjusted_Death_Rate, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Boxplot of Age-adjusted Death Rate by Sex (Synthetic Data)", x = "Sex", y = "Age-adjusted Death Rate")
```

### Boxplot of Age-adjusted Death Rate by Sex (Synthetic Data)

#### Description:

This boxplot compares the distribution of age-adjusted death rates
across different sex categories in the synthetic dataset: Both Sexes,
Female, and Male.

#### Findings:

-   The median death rate for both sexes and females is lower compared
    to males, consistent with the original data.

-   The interquartile range (IQR) for males is slightly larger,
    indicating more variability in this category.

-   There are several outliers in the male category, suggesting
    instances of unusually high death rates, which is also consistent
    with the original data.

-   The boxplot confirms that the synthetic data maintains the same
    distribution characteristics and variability as observed in the
    original dataset.

## Boxplot of `Average_Life_Expectancy_Years` by `Race`:

```{r}
ggplot(synthetic_data, aes(x = Race, y = Average_Life_Expectancy_Years, fill = Race)) +
  geom_boxplot() +
  labs(title = "Boxplot of Average Life Expectancy (Years) by Race (Synthetic Data)", x = "Race", y = "Average Life Expectancy (Years)")
```

### Boxplot of Average Life Expectancy (Years) by Race (Synthetic Data)

#### Description:

This boxplot visualizes the distribution of average life expectancy
across race categories in the synthetic dataset: All Races, Black, and
White.

#### Findings:

-   The median life expectancy is similar for All Races and White
    categories, slightly lower for Black, consistent with the original
    data.

-   The interquartile range is largest for the White category,
    indicating more variability in this group.

-   The Black category shows the lowest median life expectancy.

-   The synthetic data successfully mimics the distribution
    characteristics of the original dataset, including the central
    tendencies and variability across race categories.

# Conclusion

In this analysis, I successfully generated synthetic data based on the
original cleaned dataset provided by Sri. I then performed exploratory
data analysis (EDA) on the synthetic data, producing summary statistics,
histograms, and boxplots to compare with the original data. This
approach helps ensure the synthetic data closely mimics the
characteristics and distributions of the original data, enabling further
analysis and modeling.

# Comments on the Analysis

1.  **Data Cleaning**: I started by loading Sri's original data and
    renaming columns to avoid spaces, which synthpop requires for
    generating synthetic data correctly.

2.  **Synthetic Data Generation**: Using the `synthpop` package, I
    generated a synthetic dataset based on the original data. The
    synthetic data retains the structure and distribution of the
    original dataset.

3.  **Exploratory Data Analysis (EDA)**: I conducted a similar EDA on
    the synthetic data as Sri did on the original data, including
    summary statistics, histograms, and boxplots.

4.  **Comparison**: The EDA of the synthetic data closely mirrors the
    original data, demonstrating that the synthetic data maintains
    similar distributions and relationships among variables. This is
    crucial for ensuring that any analysis or modeling performed on the
    synthetic data would yield results similar to those obtained from
    the original data.

By generating and analyzing synthetic data, I can share and analyze
sensitive datasets without compromising privacy, ensuring that the
synthetic data provides a reliable stand-in for the original dataset.
This approach allows us to conduct meaningful analyses and derive
insights while protecting the confidentiality of the original data.
